---
title: "Curate Data in R"
author: "Brian S. Yandell"
date: "8/23/2017"
output: html_document
---

```{r knitr_options, echo=TRUE}
```

------------

This and other lessons assume users already have some experience with R, for instance through a [Data Carpentry](http://kbroman.org/datacarpentry_R_2017-01-10/) workshop. See [References](reference.md) for useful introductory and advanced material.

> ## Learning Objectives
>
> After completing this section, an individual will be able to curate data in `R`.
>
> * read, manipulate and display data summaries in concise tables
> * work with data frames using [tidyverse](http://tidyverse.org/) tools
> * use character string operations to clean data
> * create functions to collapse repeated steps into one-line "verbs"
> * save cleaned up data tables in external files

------------

## Read, Manipulate and Display Data

* [Working with data tables in R](curate/data_tables.html)
* [Examples using `for` and `apply`](curate/applyExample.html)

## Work with Data Frames using Tidyverse tools

* [sub_lesson recasting data.frame examples](curate/intro_dplyr.html)
* [Tidyverse split-apply-combine example](curate/tidyverse.html)

## Character String Operations to Clean Cata

* [characters and strings](string.Rmd)
* [regular expressions (regex)](regex.Rmd)

## Create Functions to Collapse Steps

[examples of functions](function.Rmd)

## Save Cleaned up Data to External Files

[read and write files](file.Rmd)

### Challenge

What is the value of `y` after doing the following?

```{r, eval=FALSE}
x <- 50
y <- x * 2
x <- 80
```

<!-- end challenge -->

