---
title: "Zoology Dplyr Workshop"
author: "Brian S. Yandell"
date: "11/14/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This is a 30-45 minute session on data wrangling, primarily with the `dplyr` package.
Material is drawn from the "Aggregating and analyzing data with dplyr" session that is part of the 2-day [Data Carpentry](http://www.datacarpentry.com) workshop taught on campus several times a year.

Websites for the January 2018 workshops can be found at these addresses. Registration opens for BOTH on December 11 at 5:00.  

- Data Carpentry, January 8 - 9
<https://uw-madison-aci.github.io/2018-01-08-uwmadison-dc/>

- Software Carpentry, January 10-11
<https://uw-madison-aci.github.io/2018-01-10-uwmadison-swc/>

This session draws heavily on
<http://kbroman.org/datacarpentry_R_2017-01-10/02-dplyr.html>.

## Setup

Please have the latest version of `R` (3.4.2 from <http://cran.r-project.org>) and `RStudio` (1.1.383 from <http://rstudio.com>). In addition, please install the `tidyverse`. You can do that within `RStudio` by going to the `Packages` tab of the lower right pane, click `Install`, type `tidyverse`, and click `Install` button. Or you can type the following on the console (lower left pane):

```{r eval=FALSE}
install.packages("tidyverse")
```


