---
title: "Example Rmarkdown"
output:
  pdf_document: default
  html_document:
    toc: yes
    toc_float: yes
---
#Large  header
 
Regular text, non-code. 
 
&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;&nbsp;How to indent text
 
*You can italicize things.*
**You can bold things.**
Notice how a single paragraph break shows up as a space.
 
##Slightly smaller sub-header
 
###Even smaller sub-header
 
####You get the idea
 
Including a link http://cran.mtu.edu/
 
```{r ThisLabelIsForYouItsNotShownNoSpaces}
#This all shows up in a "code" grey box
#Code will be automatically colored to highlight file names, command modifiers, etc.
myData = data.frame(replicate(10,sample(0:1,1000,rep=TRUE)))
```
 
# Next large header
 
You can include the output of some in-line code like: The dimensions of my data are `r dim(myData)`
 
And you can also include the actual in-line code like `dim(myData)`
 
## Making some figures

```{r MakeSomePlots}
hist(myData$X1)
```
 
Modifying all figures in a chunk with parameters in the chunk header
```{r AddSomePlotParameters,fig.width=3.5,fig.height=4}
hist(myData$X2)
hist(myData$X3)
```
 
## Analyzing some data

This option only shows the output, not the code

```{r echo=FALSE}
#Comments like this won't be shown either
x <- rnorm(100)
y <- 2*x + rnorm(100)
cor(x, y)
```
 
This option only shows the code, not the output

```{r results="hide"}
x <- rnorm(100)
y <- 2*x + rnorm(100)
cor(x, y)
```
 
This option evaluates the code but neither code nor output is displayed

```{r include=FALSE}
x <- rnorm(100)
y <- 2*x + rnorm(100)
cor(x, y)
```
