---
title: "sqlite"
author: "Brian S Yandell"
date: "September 1, 2016"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## SQLite with R/dplyr Introduction

This is an example using the `dplyr` package to access SQLite. It draws on lessons from the [UW-Madison Data Carpentry Workshop](http://uw-madison-aci.github.io/2016-06-01-uwmadison/). Other resources include the [Data Wrangling Cheat Sheet](https://www.rstudio.com/resources/cheatsheets/) and the [R/dplyr Databases Vignette](https://cran.rstudio.com/web/packages/dplyr/vignettes/databases.html).
For more information on SQL in general see the [Lynda.com SQL Essential Training](https://www.lynda.com/SQL-tutorials/SQL-Essential-Training/139988-2.html) or the [W3 SQL Tutorial](http://www.w3schools.com/sql/).

### Motivation

The [Databases with SQL](http://www.datacarpentry.org/sql-ecology-lesson/00-sql-introduction.html) lessons give a good motivation in general for using SQL. Here we stress a few points that will be important for Big Data projects in omic studies:

* SQL can make working with multiple related files manageable.
* SQL scales nicely to large projects, where you only want to examine a small portion of the data.
* Multiple people can work off the same SQL when properly set up.

### Learning Goals

The goal is to use R tools with SQL to reproduce what was done with the SQL tool embedded in Firefox, we’ll need to do the same basic data operations:

* select subsets of the data (rows and columns)
* group subsets of data
* do math and other calculations
* combine data across spreadsheets

### Setup

Make sure you have the package `dplyr`:

```
install.packages("dplyr")
```

Attach the package `dplyr`. 
The `suppressPackageStartupMessages()` suppresses messages.

```{r}
suppressPackageStartupMessages({
  library(dplyr)
})
```

Download the SQLite file `portal_mammals.sqlite` from the [Portal Project Teaching Database](https://figshare.com/articles/Portal_Project_Teaching_Database/1314459).

It helps to save the name for later use.

```{r}
sql_file <- "../data/portal_mammals.sqlite"
```

## Basic queries: 

### Connecting to SQLite database tables

Connect to the SQLite database with the command `src_sqlite()`. There are similar commands for other types of SQL databases.

```{r}
sql_connect <- src_sqlite(sql_file)
```

Connect to a table in the SQLite database.

```{r}
species_sql <- tbl(sql_connect, "species")
species_sql
```

The object `species_sql` is not actually a data table, even though it looks like it is. Notice for instance that the table begins and ends with

```
## Source:   query [?? x 4]
# ... with more rows
```

since the number of rows is not yet known (`??`). Typicall we refine this object to the rows and columns we want, and then `collect` it from SQL to have the data "in hand" that we want.

```{r}
species_all <- species_sql %>% collect
species_all
```

The `species` table is fairly small, but if it were huge, we might only want to work with a part of it. The SQL verbs `SELECT` and `WHERE` have comparable R/dplyr verbs `select()` and `filter()`. We use the dplyr "pipes" to make this nice and clean.

### Select columns

Select `year`, `month` and `day` only from table `surveys`.

```{r}
tbl(sql_connect, "surveys") %>%
  select(year,month,day)
```

### Distinct (unique) entries in a column.

Just get distinct `species_id`.

```{r}
tbl(sql_connect, "surveys") %>%
  distinct(species_id)
```

or `species_id` and `year`

```{r}
tbl(sql_connect, "surveys") %>%
  distinct(year,species_id)
```

### Calculated values

Calculations in `dplyr` are often done with the `mutate()` verb. The `filter()` verb (explained below) removes rows with missing `weight_kg` values.

```{r}
tbl(sql_connect, "surveys") %>%
  mutate(weight_kg = weight / 1000) %>%
  filter(!is.na(weight_kg)) %>%
  select(plot_id,species_id,weight_kg)
```

You can use `mutate()` to change column values without creating new columns. Here we change `species_id` to lower case.

```{r}
tbl(sql_connect, "surveys") %>%
  mutate(weight_kg = weight / 1000,
         species_id = tolower(species_id)) %>%
  filter(!is.na(weight_kg)) %>%
  select(plot_id,species_id,weight_kg)
```

### Filter rows

Filter on `species_id` being `DM`

```{r}
tbl(sql_connect, "surveys") %>%
  filter(species_id == "DM")
```

Filter on `species_id` being `DM` and `year` past 2000.

```{r}
tbl(sql_connect, "surveys") %>%
  filter(species_id == "DM",
         year > 2000)
```

Filter on `species_id` being one of `DM`, `DO` or `DS`.

```{r}
tbl(sql_connect, "surveys") %>%
  filter(species_id == "DM" |
           species_id == "DO" |
           species_id == "DS")
```

Same result done with the `%in%` operator.

```{r}
tbl(sql_connect, "surveys") %>%
  filter(species_id %in% c("DM","DO","DS"))
```

We can also add comments with `#` character.

```{r}
tbl(sql_connect, "surveys") %>%
  filter(species_id %in% c("DM","DO","DS"), # subset of species
         year > 2000) # recent years
```

### Order rows

One nice way to order is with the `arrange()` verb. I added a filter to eliminate any rows with missing `species_id` or `sex`.

```{r}
tbl(sql_connect, "surveys") %>%
  filter(!is.na(species_id),
         !is.na(sex)) %>%
  arrange(species_id)
```

We can arrange by more than one column:

```{r}
tbl(sql_connect, "surveys") %>%
  filter(!is.na(species_id),
         !is.na(sex)) %>%
  arrange(species_id,hindfoot_length,sex)
```

### Order of Execution

SQL clauses are written in a fixed order, but R/dplyr sorts that out for us.
We can put them in any order, provided the columns and rows we want to act on are available at that step.

Suppose we want to select `genus` and `species` but filter to only include the `taxa` labeled `"Bird"`. 

```{r}
species_genus_Bird <- tbl(sql_connect, "species") %>%
  filter(taxa == "Bird") %>%
  select(genus,species)
```

Notice that we filter first, as once we select `genus` and `species`, we no longer have the `taxa` column. This behavior is somewhat different from the SQLite commands highlighted in Data Carpentry, so watch out. The following will not work:

```
## This does not work -- why?
tbl(sql_connect, "species") %>%
  select(genus,species) %>%
  filter(taxa == "Bird")
```

## SQL Aggregation

### Count and Group By

The SQL `COUNT` and `GROUP BY` have corresponding verbs in `dplyr`. The `group_by()` verb enables us to group by variables and do various things. What is often done is to `summarize()` some aspect of variables by group.

The count verb is `n()`, and there are other useful verbs such as `row_number()`. However, these will not work properly on SQL objects, but rather on tables after we `collect` them. 

```{r}
species_all %>%
  group_by(genus) %>%
  summarize(count = n()) %>%
  arrange(desc(count))
```
### Having

The SQL `HAVING` does not have an analogous verb. However, you can get the same thing using `group_by()` and `summarize()` and `filter()`. Note that here we get the count of the number rows with missing `species_id`.

```{r}
tbl(sql_connect, "surveys") %>%
  group_by(species_id) %>%
  summarize(count = n()) %>%
  filter(count > 10)
```

### Joins and Merges

One great power of databases is to merge or join data from tables of different type. SQLite uses the `JOIN` verb (with possible prefix), while dplyr uses a variety of `*_join`. The dplyr package has a variety of ways to join data, depending on the purpose. We can only introduce the ideas here. See the dplyr cheatsheet and manual page for `join` for more information.

The dplyr join routines will join by column name and value. 

Note: There is a `merge()` verb in base R. This is not what we are examining.

Let's `collect` the subset of surveys examined above:

```{r}
surveys_2001 <- tbl(sql_connect, "surveys") %>%
  filter(species_id %in% c("DM","DO","DS"), # subset of species
         year > 2000) %>% # recent years
  collect
dim(surveys_2001)
```

To recall, we recreate `species_all`.

```{r}
species_all <- tbl(sql_connect, "species") %>%
  collect
dim(species_all)
```

```{r}
species_surveys <- full_join(species_all, surveys_2001)
dim(species_surveys)
```
Notice how the `left_join` and `right_join` are different sizes. Why is this?

```{r}
dim(left_join(species_all, surveys_2001))
dim(right_join(species_all, surveys_2001))
```

The `anti_join` will return rows of `species_all` that have no `species_id` in `surveys_2001`. Why are these not found in `surveys_2001`?

```{r}
anti_join(species_all, surveys_2001)
```

## Advanced Topics

### Other dplyr verbs for SQL Databases

There are a host of other dplyr `verbs`. See the references at the top of this document for help. The `databases` document is part of the `dplyr` package: type `databases` in the `Help` tab of Rstudio, or to go to the `Packages` tab, click on `dplyr` and then click on `User guides, package vignettes and other documentation`.

You can also use arbitrary SQL directly:

```{r}
tbl(sql_connect, build_sql(
  "SELECT species_id, COUNT(surveys.species_id)
   FROM surveys
   GROUP BY species_id
   HAVING COUNT(surveys.species_id) > 10")) %>%
  collect
```

### Big Data

With very large tables, say many thousands or millions of rows, it makes sense to `select` and `filter` before the `collect` stage, so that the collected object is manageable size and is focused on material you want. For instance, for genomic data, one typically `filter`s to a chromosome and a megabyte region, possibly `select`ing a subset of columns. With SQLite, this is very fast, and ensures that you only have in hand what you want.

That is, the basic philosophy is to clean up your data, store it somewhere, place it in a SQL database, and then collect only the parts you need when you need them. The SQLite database could be in another part of your computer, in a folder in Box, or somewhere on the internet. [There may be access issues, and time delays with some configurations.] 

Let's compare the complete `species` table with the reduced table for `Bird`.

```{r}
species_all <- tbl(sql_connect, "species") %>%
  collect
dim(species_all)
```

```{r}
species_genus_Bird <- tbl(sql_connect, "species") %>%
  filter(taxa == "Bird") %>%
  select(genus,species) %>%
  collect
dim(species_genus_Bird)
```

Remember, it is only at the `collect` step that data are collected from the SQL database. Fewer columns and rows means less space and faster subsequent computation.

*Challenge:* Why do you not need to save large data files locally? Think about several reasons why and why not to save large data files on your computer.

### Creating SQLite Database with R/dplyr

SQL has a `CREATE` verb to create a database and tables within it. Here are the basic steps to create an SQLite database in R with `dplyr`.
We illustrate with the `species_all` and `surveys_2001` objects. 
Here are the commands to create the database and copy `species_all` and `surveys_2001` in with names `species` and `surveys`, respectively.

```
my_sql_db <- src_sqlite("my_db.sqlite", create = TRUE)
copy_to(my_sql_db, species_all, "species", temporary = FALSE)
copy_to(my_sql_db, surveys_2001, "surveys", temporary = FALSE)
```

The above was not actually run. The code below is run, but has some
checks to see if things exist already.
Give a path to create an SQLite database. Here we check if the database exists, and if so, define `my_tbls` as the set of tables already in the database.

```{r}
if(!file.exists("my_db.sqlite")) {
  my_sql_db <- src_sqlite("my_db.sqlite", create = TRUE)
  my_tbls <- NULL
} else {
  my_sql_db <- src_sqlite("my_db.sqlite", create = FALSE)
  my_tbls <- collect(tbl(my_sql_db, "sqlite_stat1"))$tbl
}
```

Now `copy_to` this database the species table

```{r}
if(!("species" %in% my_tbls)) {
  copy_to(my_sql_db, species_all, "species", temporary = FALSE)
}
```

and the survey table.

```{r}
if(!("surveys" %in% my_tbls)) {
  copy_to(my_sql_db, surveys_2001, "surveys", temporary = FALSE)
}
```

The SQLite database now has three tables.

Using the option `temporary = FALSE` forces the tables to be written into the SQLite database.

```{r}
my_sql_db
```
The `sqlite_stat1` is a table to manage information about the database. You can look at it, but this one is not very interesting.

```{r}
tbl(my_sql_db, "sqlite_stat1")
```

We can check the size of the saved file within R, or by looking at the directory outside of R. This file can be saved somewhere for later use.

```{r}
file.size("my_db.sqlite")
```

### Indexing

Indexing of SQL is a way to speed up access for large database tables. 
The SQL verb is `INDEX`.
This is done in dplyr with `copy_to()` at the time the table is created.

```{r}
if(!("surveys_index" %in% my_tbls)) {
  copy_to(my_sql_db, surveys_2001, "surveys_index", temporary = FALSE,
          indexes = list(sp="species_id",yearsex=c("year","sex")))
}
```
Here we see that the indexes were actually created.

```{r}
tbl(my_sql_db,"sqlite_stat1")
```

### Dropping SQL Tables

Generally, one should take great care in deleting, removing or dropping electronic objects. Once gone, it may be gone forever. Having said that, there is an SQL `DROP` verb to drop a table from a database, or a whole database.
One can also simply remove the `*.sqlite` file from a folder.

In `dplyr`, the drop verb is not revealed to users quite so directly, for the reasons stated above. There is a command `db_drop_table()` but it is pretty arcane. Here we show how to drop the `surveys` table from the database, but we don't actually run it in this markdown document.

```
db_drop_table(my_sql_db$con,"surveys")
```

### Structure of SQL Connection

The R/dplyr SQL verbs use "lazy evaluation". That is, operations are only done when absolutely necessary. The SQL connect structure created by a call to `src_sqlite()` is basically a set of pointers into the SQL database.
It does not actually contain any of the data, but instead points to where the data are and what to do to get the requested parts at the moment of `collect`ion.

```{r}
sql_connect <- src_sqlite(sql_file)
species_sql <- tbl(sql_connect, "species")
str(species_sql)
```

The object `species_genus_Bird` before `collect`ing a set of instructions in machine-readable form to get the data from the SQL database. Look carefully to see reference to `taxa`, `genus` and `species`. 

```{r}
species_genus_Bird <- tbl(sql_connect, "species") %>%
  filter(taxa == "Bird") %>%
  select(genus,species)
str(species_genus_Bird)
```
