---
title: "apply example"
author: "Brian S. Yandell"
date: "7/5/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The purpose of the base `apply` family of functions is to SIMPLIFY loops.

## loop approach

```{r}
numSimulations <- 10000
ptm <- proc.time()
results <- NULL
for(iS in 1:numSimulations){
  temp <- rnorm(100, 0, 1)
  results <- rbind(results, 
                   c(simNumber = iS, 
                     mean = mean(temp), 
                     min = min(temp),
                     max = max(temp)))
}
(loopTime <- proc.time() - ptm)
```

## apply approach - "apply" function

```{r}
ptm <- proc.time()
simulation <- function(ID){
  temp <- rnorm(100, 0, 1)
  c(simNumber = ID, 
    mean = mean(temp), 
    min = min(temp), 
    max = max(temp))
}

results <- sapply(1:numSimulations, simulation)
results <- t(results)
(applyTime <- proc.time() - ptm)
```

But this is misleading; the bottleneck in the loop
is that the array is being continuously copied and
memory reallocated.  If instead we allocate the memory
initially, then the times become quite close.

```{r}
ptm <- proc.time()
results <- matrix(0, nrow = numSimulations, ncol = 4)
for(iS in 1:numSimulations){
  temp <- rnorm(100, 0, 1)
  results[iS,] <- c(simNumber = iS, 
                    mean = mean(temp), 
                    min = min(temp),  
                    max = max(temp))
}
(loopTime2 <- proc.time() - ptm)
```

### lapply and sapply

`lapply` takes a vector or list (incl. data.frame) and returns list.
`sapply` does the same but returns a simplified structure.

```{r}
myList <- list(a = c(1,2,3,4,5), 
               b = c(6,7,8,9,10), 
               c = c(11,12,13,14,15))
fun_item1 <- function(data1, data2){ data1 }
fun_type1 <- function(data1) { list(typeof(data1), data1) }

lapply(myList, mean)
sapply(myList, mean)

myVector <- 0:10
lapply(myVector, `^`, 2)
sapply(myVector, `^`, 2)

myDF <- data.frame(names = c("ann","bob","corinne"),
                   salary = c(15000, 25000, 150000),
                   age = c(50, 40, 60))
lapply(myDF, mean)
sapply(myDF, mean)
```

### vapply

`vapply` works on vectors, returns an array, requires FUN.VALUE

```{r}
myStats <- function(data){ 
  c(mean=mean(data), min=min(data), max=max(data))
  }
vapply(myList, myStats, 
       FUN.VALUE = c(Avg = 0, minValue = 0, maxValue = 0))
```

### apply

`apply` works on the margin of an array

```{r}
myArray <- matrix(1:12, ncol=3)
apply(myArray, 1, sum)
apply(myArray, 2, sum)
```

### tapply

`tapply` uses factors instead of margins

```{r}
myVector <- 1:12
myFactors <- rep(c("a","b"), c(5,7))
tapply(myVector, myFactors, min)
```

### mapply

`mapply` works where functions require more than one argument

```{r}
mapply(`^`, 1:4, 2:5)
```
