---
title: "R functions"
author: "Brian S. Yandell"
date: "7/5/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

A simple function has a name, possibly one or more arguments, and a body.

```{r}
calcGCPct <- function(xSeq) {
  length(which(xSeq %in% c("C","G"))) / length(xSeq)
}
```

```{r}
mySeq <- sample(c("A","C","G","T"), 1000, replace = TRUE)

calcGCPct(mySeq)
calcGCPct(mySeq[1:100])
```

Notes about arguments

- order: full names, abbreviated names, order
- assignment: = v. <-
- elipsis (`...`) for arbitrary additional arguments
- defaults

```{r}
makeVector <- function(aa, ab, ba) { c(aa, ab, ba) }
makeVector(1, 2, 3)
makeVector(1, 2, aa = 3)
makeVector(1, ab = 3,a = 2)

makeAlphanumericSample <- function( alphabet, 
                                    seqLength = 1000, 
                                    useProbabilities = NULL) {
  if(is.null(useProbabilities)) 
    useProbabilities <- rep(1/length(alphabet), length(alphabet))
  sample(alphabet, seqLength, replace=TRUE, prob=useProbabilities)
}

makeDNASample <- function(...){
  makeAlphanumericSample(c("A","C","G","T"), ...)
}

makeDNASample(seqLength = 100)
makeDNASample(seqLength = 100, 
              useProbabilities = c(0,.5,.5,0))
```

Notes about return values:

- returns last evaluated expression.
- thus only one object, but object can be a list of objects
- Hadley Wickham suggests reserving "return(expr)" for early returns.

The following returns `NULL`:

```{r}
testFunction <- function(){}
testFunction()
```

```{r}
testFunction <- function(x) 5+7
testFunction()
```

```{r}
testFunction <- function(x) { 5+7; 13 }
testFunction()
```

```{r}
testFunction <- function(x){
  if(missing(x)) 
    return(0)
  x^2
}
testFunction()
testFunction(4)
```

Notes about scope

- R looks first in the local environment for a variable
- if not found, it looks at the environment of the calling function
- can force use of Global environment by using <--

```{r}
testFunction <- function(x){ 
  y <- 5
  x + y 
}
rm(y)
```

```{r}
testFunction(2)
```

```{r eval = FALSE}
y
```

```
## Error: object 'y' not found
```

```{r}
y <- 2
testFunction(2)
y

```{r}
testFunction <- function(x) {
  v1 <- x + y
  y <- 10
  v1 + y
}
y<-2
testFunction(4)
```

```{r}
testFunction <- function(x){
  y <<- 5
  x + y
}
rm(y)
testFunction(2)
y
```
