---
title: "purrr package"
author: "Brian S. Yandell"
date: "9/18/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The `dplyr` package acts on data frames, `filter`ing rows and `select`ing columns as one would in a database. The `purrr` package acts on lists, including data frames. The cool thing is that the lists may contain rather complicated objects. There is now a [purrr cheatsheat](https://github.com/rstudio/cheatsheets/raw/master/purrr.pdf) Below is a made-up example to illustrate some features.

```{r message=FALSE}
library(purrr)
library(dplyr)
library(tidyr)
```


Let's create some fake data. Here `x` is random and `y` equals column number + `x` + noise.

```{r}
nc <- 5
nr <- 10
```

```{r}
x <- as.list(seq_len(nc))
```


```{r}
x <- as.data.frame(matrix(rnorm(nr * nc), nr, nc))
x <- as.data.frame(x)
```

```{r}
y <- t(seq_len(nc) + t(x) * seq_len(nc)) + 
  matrix(rnorm(nr * nc,, 0.001), nr, nc)
y <- as.data.frame(y)
```

Now we want to run a regress of each column of `x` on each column of `y` and report out the coefficients. This could be done in a loop, but let's instead imagine what we would do with a single column.

```{r}
lm_coef <- function(x1, y1) {
  dat <- data.frame(x = x1, y = y1)
  coef(lm(y ~ x, dat))
}
```

```{r}
lm_coef(x[,1], y[,1])
```

We want to combine `x` and `y` into one list. We first make a list of lists, then transpoe it.

```{r}
xy <- list(x = as.list(x), y = as.list(y))
str(xy)
```

```{r}
xy <- xy %>%
  transpose
str(xy)
```

```{r}
xy %>% 
  map(function(dat) coef(lm(y~x, dat)))
```

This can also be done in two steps. At the end, we organize data a bit.


```{r}
xy %>% 
  map(function(dat) lm(y~x, dat)) %>%
  map(coef) %>%
  as.data.frame %>%
  t
```

## Redo with dplyr

Of course all the above can be done readily with `dplyr` and `tidyr` using `group_by` and `do`, as shown below. However, two things are useful to consider:

* While `summarize` works for single value operations, you must use `do` for multiple value operations. It is challenging to get `do` correct, as it must return a one-row data frame.
* The `purrr` verbs `map` and `transpose` do not require the lists to be the same length or configuration. Thus, they can be used in a variety of settings where `dplyr`, working on data frames, cannot.

```{r}
xx <- x %>%
  gather(var, xval)
yy <- y %>%
  gather(var, yval)
xx$yval <- yy$yval
```


```{r}
xx %>% 
  group_by(var) %>%
  do(
    as.data.frame(
    t(
      coef(
        lm(yval ~ xval, .)))))
```

## A more complicated example

This example can be found in the [hotspot.R](https://github.com/byandell/qtl2shiny/blob/master/R/hotspot.R) file of package [qtl2shiny](https://github.com/byandell/qtl2shiny).

We start with two objects -- `chr_pos` has positions within each chromosome, `peaks` has information about what traits have peaks where on each chromosome -- that are lists by chromosome. [The `peaks` object is turned into a list by use of `split`.] We transpose a "list of lists" so that `out_chr` is also a list by chromosome, having elements from the two objects.

```{r eval = FALSE}
out_chr <- purrr::transpose(list(pos = chr_pos, 
                                   peaks = split(peaks, peaks$chr)))
```

Now we create a new object `out_peaks` by counting peaks at each position along each chromosome. The `map` function acts across the list indexed by chromosome.  Notice that the call to `peaks_type`, done for each chromosome, uses elements `x$pos` and `x$peaks` from each element of `out_chr`. Recall the `pos` and `peaks` objects have rather different shapes.

```{r eval = FALSE}
out_peaks <- purrr::map(out_chr,
                        function(x, peak_window) 
                          peaks_type(x$pos, x$peaks, peak_window),
                        peak_window)
```

The routine `peaks_type` first `split`s the `peaks` by `pheno_type`, then `map`s a function `outer_window` (not shown here) for each element of `peaks_by_type`.

```{r eval=FALSE}
peaks_by_type <- split(peaks, peaks$pheno_type)
out <- purrr::map(peaks_by_type,
                  outer_window,
                  posi, peak_window)
```

Later on, `peaks_type` uses `map` again for another summary.

The point here is not to follow the details, but to notice how `map` can be used to isolate thinking about a particular task. Then you can concentrate on solving that task without worrying about a bunch of indexing and looping.
