---
title: "regex"
author: "Brian S. Yandell"
date: "7/5/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

- <http://regexr.com>: Test out your reg expressions before you use them (specify your test expression and it will highlight in a sample document (which you can change) the matches.
- [regex base-R cheat sheet](http://www.endmemo.com/program/R/grep.php)
- [Helpful examples on grouping](http://www.regular-expressions.info/rlanguage.html)
- [Examples on subsetting matches](http://www.dummies.com/how-to/content/how-to-use-regular-expressions-in-r.html)
- [Getting strsplit in the right format with sapply](http://www.r-bloggers.com/regular-expressions-in-r-vs-rstudio/)
- [RegexBuddy](http://www.regexbuddy.com/rlanguage.html) 


Drew’s example (using `gsub`): 

Searching through these ugly column headings in data frame called `x`:

```
Raw Data (595) 1 - 0 h  
Raw Data (595) 5 - 0 h 30 min 
Raw Data (595) 7 - 1 h  
Raw Data (595) 10 - 1 h 30 min 
Raw Data (595) 13 - 2 h  
Raw Data (595) 16 - 2 h 30 min 
```

```{r}
x <- read.delim("x.txt",header=F,stringsAsFactors = FALSE)
```

I used the following `gsub()` statement (within an `apply()` function, but that’s not important here) on data frame `x`:

```{r}
gsub('\\.*(\\d*) h (\\d*).*', '\\1.0\\2',x[[1]])
```

(On a second look, it’s possible the initial `\\.*` isn’t necessary, but as always, test your regex’s out before executing them)

perl example

```{r}
gsub('\\.*(\\d*)\\sh\\s(\\d*).*', '\\1.0\\2',x[[1]], perl=T)
```

`\\s` means “space” which would also match `\\t` and `\\n`  
`\\d{2}` means 2 digits  
`[5-9]  [a-z]  \W  \D`

And it catches the following substrings (highlighted in cyan):

```
Raw Data (595) 1 - 0 h  
Raw Data (595) 5 - 0 h 30 min 
Raw Data (595) 7 - 1 h  
Raw Data (595) 10 - 1 h 30 min 
Raw Data (595) 13 - 2 h  
Raw Data (595) 16 - 2 h 30 min 
```

```{r}
d <- data.frame(c("Raw Data (595) 1 - 0 h",  
  "Raw Data (595) 5 - 0 h 30 min",
  "Raw Data (595) 7 - 1 h",
  "Raw Data (595) 10 - 1 h 30 min",
  "Raw Data (595) 13 - 2 h",
  "Raw Data (595) 16 - 2 h 30 min" ))
class(d)
dim(d)
dl <- as.list(d)
```

```
[1] 6 1
[1] "data.frame"
```
Perl =T version

```{r}
gsub('\\.*(\\d*)\\sh\\s(\\d*).*', '\\1.0\\2',d, perl=T)
gsub('\\.*(\\d*)\\sh\\s(\\d*).*', '\\1.0\\2',dl, perl=T)
```

```
[1] "c(1, 5, 6, 2, 3, 4)"
[1] "c(1, 5, 6, 2, 3, 4)"
```

Non Perl version:

```{r}
gsub('\\.*(\\d*) h (\\d*).*', '\\1.0\\2',d)
gsub('\\.*(\\d*) h (\\d*).*', '\\1.0\\2',dl)
```

```
[1] "c(1, 5, 6, 2, 3, 4)"
[1] "c(1, 5, 6, 2, 3, 4)"
```

Drew Doering used this in the context of converting time values in the format “X h Y min” to a decimal hour format, to facilitate downstream analysis/plotting. You can see that specific command in-context here on line 49: <https://github.com/dtdoering/grofitr/blob/b70f82755e83ad27b4ebe07c9e78b3cc36159351/findRates.R>

However, playing with this in my own instance of RStudio, it doesn’t seem to be doing what I wanted it to do with that specific command. Here is a reproducible example you can use to replace that last command at the bottom:

```{r}
timePoints <- c("Raw Data (595) 1 - 0 h",
                "Raw Data (595) 5 - 0 h 30 min",
                "Raw Data (595) 7 - 1 h", 
                "Raw Data (595) 10 - 1 h 30 min", 
                "Raw Data (595) 13 - 2 h", 
                "Raw Data (595) 16 - 2 h 30 min",
                "Raw Data (595) 50 - 21 h 30 min")

sapply(timePoints,
       function(x) gsub('.* - (\\d*) h( (\\d*).*)?', 
                        '\\1.0\\3',
                        x, perl = F), 
       USE.NAMES = FALSE)
```

Drew would then go on to do some string operations and arithmetic to divide the right side of the decimal by 60.
