---
title: "Tidyverse split-apply-combine example"
author: "Brian S. Yandell"
date: "8/23/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
surveys <- read.csv("http://kbroman.org/datacarp/portal_data_joined.csv",
                    stringsAsFactors = FALSE)
```

### Split-apply-combine using dplyr

```{r, message = FALSE}
library(dplyr)    ## load the package
```

```{r}
surveys %>%
  filter(taxa == "Rodent",
         !is.na(weight)) %>%
  group_by(sex,genus) %>%
  summarize(mean_weight = mean(weight))
```

### Split-apply-combine using dplyr and tidyr

```{r}
library(tidyr)
```

Here we use `spread()` from the `tidyr` package. This gives a more compact, if wider,
table summary.

```{r}
surveys %>%
  filter(taxa == "Rodent",
         !is.na(weight)) %>%
  group_by(sex,genus) %>%
  summarize(mean_weight = mean(weight)) %>%
  spread(sex, mean_weight)
```

Here is a simpler tidy example to just count how many records with weigth by genus and sex.

```{r}
surveys %>%
  filter(taxa == "Rodent",
         !is.na(weight)) %>%
  count(sex, genus) %>%
  spread(sex, n)
```

### Split-apply-combine using split, purrr and tidyr

The `purrr` package works on lists (which includes data frames), giving great generality to what objects might be in the lists. Below we show two examples.

```{r}
library(purrr)
```

In the first example, for each genus, we fit a linear model with `lm()` and extract the `"r.squared"` element from the `summary()` of the fit. Note the use of `split()` to split the data frame into a list of data frames, one per genus. The `map()` function from `purrr` returns a list, while the `map_dbl()` function returns a vector.

```{r}
surveys %>%
  filter(taxa == "Rodent",
         !is.na(weight)) %>%
  select(genus,weight,sex) %>%
  split(.$genus) %>%
  map(~ lm(weight ~ sex, data=.)) %>%
  map(summary) %>%
  map_dbl("r.squared")
```

Note that for `Spermophilus`, the $R^2$ of 1 reflects that there are only two records, one male and one female (see table above in `tidyr` section).

There were three calls to `purrr` functions. The first `map(~ lm())` call creates a list of `"lm"` objects; the second `map(summary)` call creates a list of `"summary.lm"` objects; the third `map_dbl()` creates a vector of double-precision values.

The following more complicated example takes the `coef()` of the `lm()` fits to get the estimates of coefficients. These coefficients have to be converted into a data frame,
and then the data frames are bound together using `bind_rows()`.

```{r}
surveys %>%
  filter(taxa == "Rodent",
         !is.na(weight)) %>%
  select(genus,weight,sex) %>%
  split(list(.$genus)) %>%
  map(~ lm(weight ~ sex, data=.)) %>%
  map(coef) %>%
  map(function(x) data.frame(level = names(x), estimate = x,
                             stringsAsFactors = FALSE)) %>%
  bind_rows(.id = "genus")
```

The following is the same idea as above, but using `summary()` rather than `coef()`. We also `spread()` the data for a more compact table.

```{r}
surveys %>%
  filter(taxa == "Rodent",
         !is.na(weight)) %>%
  select(genus,weight,sex) %>%
  split(list(.$genus)) %>%
  map(~ lm(weight ~ sex, data=.)) %>%
  map(summary) %>%
  map(function(x) {
    out <- as.data.frame(x$coefficients[,1, drop = FALSE])
    out$level <- row.names(out)
    out[, 2:1]
    }) %>%
  bind_rows(.id = "genus") %>%
  spread(level, Estimate)
```

